---
title: "Coursera , Peer-graded Assignment: Course Project 1"
author: "Andrew"
date: "2024-09-16"
output: html_document
---

## 01. To Download the zipped data file and unzip it
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory
setwd("D:/Developer Room/CourseraDataScienceUsingR/reproducible research/RepData_PeerAssessment1")

# set destination filename for the zip file download
destinationFileName <- "repdata-data-activity.zip"

#set the download url and download the file
getFileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(getFileUrl,"repdata-data-activity.zip")

#check if csv file is not unzipped, then unzip
csv_filename <- "activity.csv"
if(!file.exists(csv_filename)){
  unzip(destinationFileName)
}


# Read the CSV file
exercise_data <- read.csv(csv_filename)
```


## 02. Transform that date column
````{r}
#transform that date column
exercise_data$date <- as.Date(exercise_data$date, "%Y-%m-%d")
````


## 03. To compute the total and mean number of steps taken per day
````{r}
aggregated_steps_data <- aggregate(steps ~ date, exercise_data, FUN = sum)
````

## 04. Histogram plot of total daily steps
````{r}
library(ggplot2)
g <- ggplot (aggregated_steps_data, aes (x = steps))
g + geom_histogram(fill = "blue", binwidth = 1000) +
  labs(title = " Histogram of Steps Taken Each Day ", x = "Daily Steps", y = "Frequency")
````

## 05. To calculate the mean and median daily steps
````{r}
#compute the mean daily steps
mean_dailySteps <- round(mean(aggregated_steps_data$steps, na.rm=TRUE),2)
mean_dailySteps
#compute the median daily steps
median_dailySteps <- median(aggregated_steps_data$steps, na.rm=TRUE)
median_dailySteps
````
* Mean daily steps is : `r mean_dailySteps`
* Median daily steps is :  `r median_dailySteps`

## 06. Time series plot of the average number of steps taken
````{r}
# compute average number of steps per each 5-min time interval
per_interval_steps <- aggregate(steps ~ interval,  exercise_data, mean)

#  time series plot of average number of steps per interval
h <- ggplot (per_interval_steps, aes(x=interval, y=steps))
h + geom_line()+ labs(title = " Time Series Plot of Average Steps per Interval", 
                      x = "Interval", y = "Average Steps for each day")

````


## Compute  maximum number of steps across the days
````{r}
# Maximum steps by interval
maxSteps_Interval <- per_interval_steps[which.max(per_interval_steps$steps), ] 
maxSteps_Interval
````
* Most Steps occurred at: `r maxSteps_Interval`

## Getting information on Missing Steps Values, having NAs
````{r}
# Compute total number of missing values in the steps column in data set 
total_MissingValue_count <- nrow(exercise_data[is.na(exercise_data$steps),])
total_MissingValue_count
````
* Number of missing values are: `r total_MissingValue_count`

## Resolving missing values steps (NAs)
#### Substitute the missing values (steps) with the average steps based on both the 5-minute interval and the day 

````{r}
# Create a variable/column with weekdays name
exercise_data$day <- weekdays(as.Date(exercise_data$date))


````

````{r}
library(dplyr)
# Create a variable/column with weekdays name
exercise_data$day <- weekdays(as.Date(exercise_data$date))

# create average number of steps per 5-min interval and day
averageSteps <- aggregate(steps ~ interval + day, exercise_data, mean) %>%
  mutate( steps = as.integer( round(steps,0))
        ) %>%
  arrange(steps,interval,day) %>%
  distinct(interval,day, .keep_all = T)
#
# filter out the data with only missing steps values
#take out the NAs
na_exercise_data <- exercise_data %>%
  mutate(steps = as.integer(steps)) %>%
  filter(is.na(steps)) %>%
  select(-steps) #drop the steps variable since it is all NAs

#merge to replace the missing steps with the computed averages from averageSteps
newdataMerge <- merge(na_exercise_data, averageSteps, by=c("interval", "day"))

# filter out the data with none missing steps values
hasNo_Missing_exercise_data <- exercise_data %>%
  filter(!is.na(steps))

#append the rows to get back to the original number of observations
final_exercise_data_cleaned <- rbind(hasNo_Missing_exercise_data,newdataMerge)
````

## Histogram of the total number of steps taken each day after missing values are imputed
````{r}
# Calculate the total steps per day on the merged data
dailySteps_cleanedData <- aggregate(steps ~ date, final_exercise_data_cleaned, FUN = sum)

# Create the histogram
g1 <- ggplot (dailySteps_cleanedData, aes (x = steps))
g1 + geom_histogram(fill = "green", binwidth = 1000) +
    labs(title = " Histogram of Steps Taken Each Day ", x = "Steps", y = "Frequency")
````

### To see if there is a change in the mean and median steps after cleaning the data
````{r}
# Calculate the total steps per day on the merged data
# Maximum steps by interval
new_maxSteps_Interval <- dailySteps_cleanedData[which.max(dailySteps_cleanedData$steps), ] 
new_maxSteps_Interval

#compute the mean daily steps
new_mean_dailySteps <- round(mean(dailySteps_cleanedData$steps, na.rm=TRUE),2)
new_mean_dailySteps
#compute the median daily steps
new_median_dailySteps <- median(dailySteps_cleanedData$steps, na.rm=TRUE)
new_median_dailySteps

````
#### new Most Steps: `r new_maxSteps_Interval`
#### new Mean daily steps is : `r new_mean_dailySteps`
#### new Median daily steps is :  `r new_median_dailySteps`
*old max steps was 206.1698, is now 21194
*old mean steps of 10766.19 is now  10821.1
*old median steps which was 10765 has now become 11015
 All have a significant increase when NAs were resolved

## Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
````{r}
#set a new column to indicate if the day is a weekday or on weekend
final_exercise_data_cleaned$week_type <- ifelse(final_exercise_data_cleaned$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

# create table with average steps per time interval across weekday days or weekend days
newPlot_weekday_Weekend <- aggregate(steps ~ interval+week_type, final_exercise_data_cleaned, FUN = mean)
# Make the panel plot
j <- ggplot (final_exercise_data_cleaned, aes(x=interval, y=steps))
j + geom_line()+ labs(title = " Time Series Plot of Average Steps per Interval: weekdays vs. weekends", x = "Interval", y = "Average Number of Steps") + facet_grid(week_type ~ .)
````
